---
title: "PCA.RRBS"
author: "Sarah Mueller"
date: "2025-Final"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
### Libraries

listOfPackages <- c("ggplot2","reshape2","here", "ggforce", "matrixStats","viridis","readr","vegan","factoextra","readxl", "tidyverse","dplyr", "plotly", "GGally","ggpubr","ggrepel", "missMDA")

for (i in listOfPackages){
  if(! i %in% installed.packages()){
    install.packages(i, dependencies = TRUE)
  }
  require(i,character.only=TRUE)
}

```

## Set up data for PCA

First, we want to set some variables depending on the samples we want to use in the PCA, etc. change these before running the analysis

```{r,warning=FALSE}
# define input variables -------
#sample information (tab delimited file with Sample ID, population and subspecies)
pop <- read.table("/dss/dsshome1/lxc0B/ra52qed/gitlab/swallow.projects/scripts/0_sample.lists/sample.names.85.meth.sub.pop.txt", header=T)
#any other metadata to be included
info <- read_xlsx("/dss/dsshome1/lxc0B/ra52qed/gitlab/swallow.projects/scripts/0_sample.lists/NCBI_info_updated_20220727.xlsx")
#number of samples
N = 85

#color palettes for plotting
color_sub <- c( "#39ad48", #E
                "#00316e", #G
                "#d30000", #R
                #"#9716a8", #RG
                #"#FFA500", #RT
                "#be6400", #S
                "#f4d000", #T
                "#622a0f", #TV
                "grey") #NA/unknown
```

## Load in methylation data

Here, we can also specify a subset of samples if there are individuals we want to exclude

```{r, echo=FALSE, message=FALSE, warning=FALSE}
##load in methylkit object
prop.168.20miss.merged <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/0_Pre-processing/4_methyl_calling/0_all_sites/168.20miss/prop.168.merged.20miss.rmlowv.rmout.txt", header = T)
```

```{r message=FALSE, warning=FALSE}
#Create a cleaned dataframe without hybrid individuals
prop.df <- prop.168.20miss.merged

prop.df <- prop.df %>%
  dplyr::select("site.id", contains("_R_") | contains("_G_") | contains("_T_") | contains("_S_") | contains("_TV_") | contains("_E_"))

prop.df <- prop.df %>%
  mutate(chr_pos = site.id) %>%
  separate(chr_pos, into = c("chr", "pos"))

prop.df <- prop.df %>%
  filter(chr == "chr11") %>% # First, filter for chromosome 11
  filter(
    (as.numeric(pos) >= 1.8e7 & as.numeric(pos) <= 1.83e7) | # Range 1
    (as.numeric(pos) >= 1.94e7 & as.numeric(pos) <= 1.965e7) | # Range 2
    (as.numeric(pos) >= 2e7) # Range 3 (from 2e7 onwards)
  )

#load in linked methylation loci
#meth_linked <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/3_Linkage/1_Rall/4_chr11_inv_meth.txt", header=T)

#prop.df <- prop.df %>%
#  filter(pos %in% meth_linked$bp_meth) 

#prop.df <- prop.df %>%
#  dplyr::filter(as.numeric(pos) > 1.9e7) %>%
#  dplyr::filter(as.numeric(pos) < 2e7)

prop.df <- prop.df[,1:86]

prop.df <- prop.df %>%
  mutate(row_mean = rowMeans(select(., where(is.numeric)), na.rm = TRUE)) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), row_mean, .))) %>%
  select(-row_mean)


```

## Run the PCA

```{r, include=FALSE, message=FALSE, warning=FALSE}
#Remove missing data and grab site.ids for loadings
prop.df$na <- apply(prop.df, 1, function(row) any(is.na(row)))
prop_clean <- prop.df[prop.df$na == FALSE, ]
prop_clean_siteid <- as.data.frame(prop_clean[,1])
prop_clean <- prop_clean[,c(2:86)]

#Make sure all are numeric
prop_clean[] <- lapply(prop_clean, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.character(x))
  } else {
    x
  }
})

# run a PCA on the sample matrix -----
prcomp.out <- prcomp(t(prop_clean), center = T)
summary(prcomp.out)

#Get eigen values and vectors -----
pca <- as.data.frame(prcomp.out$x)
eigenval <- ((apply(prcomp.out$x, 2, var))/sum(apply(prcomp.out$x, 2, var)))*100

# Clean-up data for plotting -------
#Assign populations and sample names
pca <- cbind(pop$V1, pca)
names(pca)[1] <- "ind"
# join pop data to pca results
pca <- dplyr::left_join(pca, pop, by = c('ind' = 'V1'))

# add any other metadata to columns
info$sample_name=as.character(info$sample_name)
info=info %>% dplyr::slice(match(pop$V1, info$sample_name))

pca <- left_join(pca, info, by=c("ind" = "sample_name"))

```

## Plotting PCA axes

Here, we can use a function that plots the PCA so we can plot several at once

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# def functions --------
plot.pca <- function(data, X, Y, colors, plot_title) {
  p <- ggplot(data = data, aes_string(x = paste0(X, " * -1"), y = Y, fill = "sub", color = "sub", label = "ind", shape = "pop")) +
    geom_point(size = 4) +
    theme_bw() +
    scale_fill_manual(values = colors, na.translate = FALSE) +
    scale_color_manual(values = c("grey33", "#00316e", "grey33", "grey33", "grey33", "grey33", "grey33", "grey33"), na.translate = FALSE) +
    # scale_shape_manual(values=c(3,21,7,22,8,23:25))+
    scale_shape_manual(values = c(3, 21:25, 1, 2, 5, 6), na.translate = FALSE) +
    ylab(paste0(Y, " (", signif(pve$pve[pve$PC == Y], 3), "%)")) +
    xlab(paste0(X, " (", signif(pve$pve[pve$PC == X], 3), "%)")) +
    # ylim(-0.3,0.2) +
    labs(fill = "", shape = "") +
    ggtitle(plot_title) +
    theme(
      plot.title = element_text(hjust = 0.5), # Center the title
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "bottom", # Move legend to the bottom
      legend.title = element_blank(),
      legend.box = "vertical",
      legend.margin = margin()
    ) +
    guides(
      color = FALSE,
      fill = guide_legend(override.aes = list(shape = 21), nrow = 1, byrow = TRUE),
      shape = guide_legend(nrow = 1, byrow = TRUE)
    )
  
  # Add labels for points above 1.5 in PC1 (assuming X is PC1)
  labeled_data <- data %>%
    filter(!!sym(X) < 2.5)

  #p <- p +
  #  geom_text(data = labeled_data, aes(x = !!sym(X), y = !!sym(Y), label = ind),
  #    hjust = 0.8, vjust = 0.4, size = 3
  #  )


  return(p)
}

##this looks at the eigenvalues and the percent of variance explained by each PC axis
plot.pve <- function(pve){
   pve.p <- ggplot(data=pve, aes(x=PC, y=pve)) +
    geom_bar(stat="identity", fill="#418979") + 
    theme_bw()  +
    ylab("Explained variation (%)") +
    xlab("Principal Component") +
    ggtitle("Scree plot - PCA ")+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  
  return(pve.p)
}

```

### Cumulative variance of components

Here we want to plot the amount of variance described by each PC axis. We do this by using the eigenvalues and converting this to a percentage of variance.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#Plot cumulative variance of components
pve <- data.frame(PC = paste0("PC", 1:(ncol(pca)-(ncol(info)+2))), pve = eigenval/sum(eigenval)*100)
pve$PC <- factor(pve$PC, levels=unique(pve$PC))
ggplotly(plot.pve(pve))
```


### Plot PCA

We can make plots looking at the first 3 axes of the PCA.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Plot first 3 PCA combos 
p1 <- plot.pca(pca, "PC1","PC2", color_sub, "RRBS")
p2 <- plot.pca(pca, "PC1","PC3", color_sub, "RRBS")
p3 <- plot.pca(pca, "PC2","PC3", color_sub, "RRBS")
p4 <- plot.pca(pca, "PC3","PC4", color_sub, "RRBS")
p5 <- plot.pca(pca, "PC2","PC4", color_sub, "RRBS")
p6 <- plot.pca(pca, "PC4","PC5", color_sub, "RRBS")

p1
p2
p3
p4
p5
p6

ggsave("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/1_Population_Structure/1_PCA/Supp.Fig.LD.inversion.png", plot = p1, width = 5, height = 5, dpi = 600)

```
#PCA Loadings

Look at which sites contribute most to which axes

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#look at loadings
loadings <- as.data.frame(prcomp.out$rotation)
#plot
dnam_sites <- rownames(prcomp.out$rotation)

#get site.id column
prop.load <- prop_clean
prop.load <- cbind(prop_clean_siteid, prop.load)

df_loadings <- data.frame(dnam = prop.load$`prop_clean[, 1]`, PC1_loadings = loadings$PC1, PC2_loadings = loadings$PC2, PC3_loadings = loadings$PC3)

#filter the loadings to only look at the top 1%
df_loadings$PC1_loadings <- abs(df_loadings$PC1_loadings) 
df_loadings$PC2_loadings <- abs(df_loadings$PC2_loadings) 
df_loadings$PC3_loadings <- abs(df_loadings$PC3_loadings) 

#filter for only top loadings
filtered_loadings <- df_loadings %>%
  filter(PC1_loadings >= quantile(PC1_loadings, 0.999))

ggplot(filtered_loadings, aes(x = dnam, y = PC1_loadings)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "DNAm Site", y = "PC1 Loadings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("PC1 Loadings")

#filter for only top loadings
filtered_loadings <- df_loadings %>%
  filter(PC2_loadings >= quantile(PC2_loadings, 0.999))

ggplot(filtered_loadings, aes(x = dnam, y = PC2_loadings)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "DNAm Site", y = "PC1 Loadings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("PC2 Loadings")

#filter for only top loadings
filtered_loadings <- df_loadings %>%
  filter(PC3_loadings >= quantile(PC3_loadings, 0.999))

ggplot(filtered_loadings, aes(x = dnam, y = PC3_loadings)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(x = "DNAm Site", y = "PC1 Loadings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("PC3 Loadings")

```

