---
title: "Hybrid Models (RT)"
author: "Sarah Mueller"
date: "2025-06-16"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
### Libraries

listOfPackages <- c("ggplot2","reshape2","here", "ggforce", "matrixStats","viridis","readr","vegan","factoextra","readxl", "tidyverse","dplyr", "plotly", "GGally","ggpubr","ggrepel", "data.table", "missMDA", "gghalves", "ggdist", "DHARMa", "betareg", "lmtest", "car")

for (i in listOfPackages){
  if(! i %in% installed.packages()){
    install.packages(i, dependencies = TRUE)
  }
  require(i,character.only=TRUE)
}

```

##Load genome information.

```{r}
category <- read.delim("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/0_Pre-processing/4_methyl_calling/1_CpG_islands/168.20miss/homer.cpgi.prop.168.merged.20miss.rmlowv.rmout.txt", header=T)

#separate chr and pos columns
category <- category %>%
  separate(loci.id, into = c("chr", "pos"), sep = "_", remove = FALSE) %>%
  distinct(chr, pos, .keep_all = TRUE) %>%
  mutate(pos = as.numeric(pos))

#get rid of leading and trailing white spaces
category$category <- trimws(category$category)

#specify region
category <- category %>%
  mutate(region = sub("_.*", "", category))

info <- read_xlsx("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/0_metadata/hirundo_WGS_sampling_metadata_10.10.22_sm.xlsx")

rename_file <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/vcf/Safran_lab/invariant_vcf/0_vcf_sample_rename.txt", header=F, sep= "\t")
colnames(rename_file) <- c("id.safran", "id.wolf")

info <- info %>%
  filter(`project sample` %in% rename_file$id.safran)

info <- left_join(rename_file, info, by = c("id.safran" = "project sample"))

rt_admix <- info %>%
  filter(grepl("_T_|_R_|_RT_", id.wolf))


##grab local ancetry information
local_anc <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/2_hybrid_index/1_introgress/1_RT_local_ancestry_HI.txt", header=T)

global.anc <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/2_hybrid_index/1_introgress/1_RT_global_ancestry_informative_HI.txt", header=T)

local_SNP <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/2_hybrid_index/1_introgress/1_RT_local_nearest_SNP_maf1.txt", header=T)

```

## Load PST information for the parental population comparisons for the hybrid zone

Looking at highest PST differences between RGT and looking at these sites in hybrids

```{r, message=FALSE, warning=FALSE}
tukey_results <- read.delim("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/4_RGT_Hybrids/3_methylation_models/RT_hyb/RT_diff/0_RT_parental_hybrid_tukey_results.txt", header=T)

```

##Load methylation data
```{r}
prop.hyb <- read.delim("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/0_Pre-processing/4_methyl_calling/1_CpG_islands/168.20miss/homer.cpgi.prop.168.merged.20miss.rmlowv.rmout.txt", header=T)

```

##Grab significant loci for the RG comparison
```{r}
#keep only revelvant info from category
category <- category %>% dplyr::select(region.id, n_cpgs, Annotation, category, region)
##Combine the category w/ selected loci
cpg_loci <- left_join(tukey_results, category, by= c("site.id" = "region.id"))

#Combine this with methylation data
cpg_loci <- left_join(cpg_loci, prop.hyb, by = c("site.id" = "region.id"))

cpg_loci <- cpg_loci %>%
  dplyr::select(!contains("DE_") & !contains("CH_") & !contains("FI_") & !contains("_G_") & !contains("_RG_")  & !contains("_S_") & !contains("_TV_") & !contains("_E_") & !starts_with("RU_"))

# Remove "ME_" from the start of any column name in pst_rg
colnames(cpg_loci) <- gsub("^ME_", "", colnames(cpg_loci))
```

##Model Testing

Essentially, I want to determine if cis- or trans- acting (or environmental/spatial) components influence methylation proportion at sites significantly different between the three groups (parent1, hybrids, parent2). Therefore, I want to run a model that take methylation prop ~ genome-wide HI + local HI (60kb surrounding focal region) + nearest_SNP (maf >0.1)

```{r}
##Format Data
#I need all individuals in rows ad then any defining variables in columns
#I will start with the response variable aka methylation
meth.loci <- cpg_loci %>% dplyr::select(contains("_ADL_"))
rownames(meth.loci) <- cpg_loci$site.id
meth.loci <- as.data.frame(t(meth.loci))
meth.loci$id <- rownames(meth.loci)

#Next the global admixture proportions and longitude/latitude data
rt_admix <- rt_admix %>% dplyr::select(id.wolf, latitude, longitude, sex, proportion_rustica_ancestry)

#Global HI
global_HI <- as.data.frame((global.anc))
global_HI$id <- global_HI$sample

#Next local HI
local_anc_for <- as.data.frame(t(local_anc))
local_anc_for$id <- rownames(local_anc_for)

#Local nearest SNP
#check distances
local_SNP <- local_SNP %>%
  separate(site_id, into = c("chr", "start", "end"), remove = F) %>%
  mutate(distance = as.numeric(pos) - as.numeric(start))
summary(abs(local_SNP$distance))

#format correctly
local_SNP_for <- as.data.frame(t(local_SNP[,6:65]))
colnames(local_SNP_for) <- paste0(local_SNP$site_id, ".snp")
local_SNP_for$id <- rownames(local_SNP_for)

#Combine data
total_data <- left_join(meth.loci, rt_admix, by = c("id" = "id.wolf"))
total_data <- left_join(total_data, local_anc_for, by = "id",  suffix = c(".meth", ".local"))
total_data <- total_data %>%
  rename_with(
    ~ ifelse(
      str_starts(.x, "chr") & !str_detect(.x, "\\.(meth|local)$"),
      paste0(.x, ".meth"),
      .x
    )
  )

total_data <- left_join(total_data, local_SNP_for, by = "id")
total_data$sex <- gsub("m\\?", "m", total_data$sex)
total_data <- left_join(total_data, global_HI, by = "id")

total_data <- total_data %>% filter(grepl("_RT_",id))
```

##Use quasi-binomial model to loop through all loci and create a table of summary statistics


```{r}
# --- 0. Prepare data (pre-checks) ---
# Ensure all columns are numeric.
# no values are outside (0,1)
RT.model.loci <- total_data %>%
  mutate(across(starts_with("chr"), ~ as.numeric(as.character(.))))

# Ensure predictors are numeric
RT.model.loci$longitude <- as.numeric(RT.model.loci$longitude)
RT.model.loci$proportion_rustica_ancestry <- as.numeric(RT.model.loci$proportion_rustica_ancestry)
RT.model.loci$all_vcf <- as.numeric(RT.model.loci$h)

# --- 1. Identify site.id columns ---
chr_columns <- names(RT.model.loci)[endsWith(names(RT.model.loci), ".meth")]

# Initialize an empty list to store results for each model
all_model_results <- list()

# --- 2. Loop through each site.id column ---
for (response_col in chr_columns) {

  cat("Processing column:", response_col, "\n") # Print progress

  # Extract the base name by removing the ".meth" suffix
  base_name <- sub("\\.meth$", "", response_col)

  # Construct the local predictor name
  local_predictor <- paste0(base_name, ".local")
  snp_predictor <- paste0(base_name, ".snp")

  # Check if the local predictor exists in the dataset
  if (local_predictor %in% colnames(RT.model.loci)) {
    formula_str <- paste(response_col, "~ all_vcf +", local_predictor, "+", snp_predictor)
  } else {
    formula_str <- paste(response_col, "~ all_vcf +", snp_predictor)
  }

  # Construct the formula
  fml <- as.formula(formula_str)

  # Fit the GLM with tryCatch
  model_fit <- tryCatch({
    glm(fml, family = quasibinomial(link = "logit"), data = RT.model.loci)
  }, error = function(e) {
    message(paste("Error fitting model for", response_col, ":", e$message))
    return(NULL)
  })

  # Proceed only if the model fitted successfully
  if (!is.null(model_fit)) {
    model_summary <- tidy(model_fit) %>%
      mutate(response_variable = response_col) %>%
      dplyr::select(response_variable, term, estimate, std.error, statistic, p.value)

    overall_model_metrics <- glance(model_fit) %>%
      mutate(response_variable = response_col) %>%
      dplyr::select(response_variable, deviance, df.residual, AIC, BIC, nobs)

    # Store both in the list
    all_model_results[[response_col]] <- list(
      coefficients = model_summary,
      overall = overall_model_metrics
    )
  }
}

# --- 3. Combine results into a single data frame ---
# Combine coefficient summaries from all models
glm.full.output_coefficients <- bind_rows(
  lapply(all_model_results, function(x) x$coefficients)
)

# Combine overall model metrics from all models
glm.full.output_overall_metrics <- bind_rows(
  lapply(all_model_results, function(x) x$overall)
)

# --- 4. Display or save the results ---
print("--- Coefficients Summary ---")
print(glm.full.output_coefficients)

print("\n--- Overall Model Metrics Summary ---")
print(glm.full.output_overall_metrics)

# Join coefficients with overall metrics for a single comprehensive table per site:
glm.full.output <- left_join(glm.full.output_coefficients, glm.full.output_overall_metrics, by = "response_variable")
print("\n--- Combined GLM Full Output (Coefficients + Overall Metrics) ---")
print(glm.full.output)

##Calculate dispersion parameter
glm.full.output <- glm.full.output %>%
  mutate(dispersion_parameter = deviance / df.residual)

#If  >1, it indicates overdispersion,  <1 it indicates underdispersion,  ≈1, the variance structure is similar to a standard binomial.

hist(glm.full.output$p.value)
sum(glm.full.output$p.value < 0.05, rm.na=T)

glm.full.output.admix <- glm.full.output %>%
  filter(term == "all_vcf")

hist(glm.full.output.admix$p.value)
sum(glm.full.output.admix$p.value < 0.05)
nrow(glm.full.output.admix)

glm.full.output.admix.sig <- glm.full.output.admix %>% filter(p.value < 0.05)

glm.full.output.local <- glm.full.output %>%
  filter(str_ends(term, "local"))

hist(glm.full.output.local$p.value)
glm.full.output.local <- glm.full.output.local %>% filter(!is.na(p.value))
sum(glm.full.output.local$p.value < 0.05)
nrow(glm.full.output.local)

glm.full.output.snp <- glm.full.output %>%
  filter(str_ends(term, "snp"))

hist(glm.full.output.snp$p.value)
glm.full.output.snp <- glm.full.output.snp %>% filter(!is.na(p.value))
sum(glm.full.output.snp$p.value < 0.05)
nrow(glm.full.output.snp)

glm.full.output.snp.sig <- glm.full.output.snp %>% filter(p.value < 0.05)

#Total Model
glm.full.output.t <- glm.full.output %>%
  filter(term == "(Intercept)")

hist(glm.full.output.t$p.value)
sum(glm.full.output.t$p.value < 0.05)
nrow(glm.full.output.t)


###Summarize terms

glm_summary <- glm.full.output %>%
  # Standardize term names
  mutate(term = if_else(str_ends(term, ".local"), "local_ancestry", term)) %>%
  mutate(term = if_else(str_ends(term, ".snp"), "nearest_SNP", term)) %>%
  # Remove intercept if you want to summarize only predictors
  filter(term != "(Intercept)") %>%
  group_by(term) %>%
  summarise(
    n_loci         = n(),                             # total number of loci this term appears in
    mean_estimate  = mean(estimate, na.rm = TRUE),
    median_estimate = median(estimate, na.rm = TRUE),
    sd_estimate    = sd(estimate, na.rm = TRUE),
    n_significant  = sum(p.value < 0.05, na.rm = TRUE), # number of loci with significant effect
    prop_significant = n_significant / n_loci           # proportion of loci significant
  ) %>%
  arrange(desc(prop_significant))

glm_summary
```

save results
```{r}
write.table(glm_summary, "/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/4_RGT_Hybrids/3_methylation_models/RT_hyb/RT_models/1_RT_model_output_summary.txt", quote=F)
```

Which sites are significant 

```{r}

glm_join <- left_join(glm.full.output.admix, glm.full.output.local, by = "response_variable")
glm_join <- left_join(glm_join, glm.full.output.snp, by = "response_variable")
glm_join <- glm_join %>%
            separate(response_variable, into = c("site.id", "extra"), sep = "\\.")

glm_join <- left_join(glm_join, cpg_loci, by = "site.id")

glm.join.sig.admix <- glm_join %>% filter(p.value.x < 0.05)

glm.join.sig.local <- glm_join %>% filter(p.value.y < 0.05)

glm.join.sig.snp <- glm_join %>% filter(p.value < 0.05)

glm.output.sig <- rbind(glm.join.sig.admix, glm.join.sig.local, glm.join.sig.snp)
```


```{r}
write.table(glm.output.sig, "/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/4_RGT_Hybrids/3_methylation_models/RT_hyb/RT_models/1_RT_model_output_sig.txt", sep = "\t", quote=F, row.names=F)
```

Look at distances of nearest SNPs 

```{r}
glm.full.output.snp.sig <- glm.full.output.snp.sig %>%
  mutate(term = str_remove(term, "\\.snp$"))

glm.distances.sig <- left_join(glm.full.output.snp.sig, local_SNP, by = c("term" = "site_id"))

summary(abs(glm.distances.sig$distance))

```

```{r}
#save results
write.table(glm.distances.sig, "/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/4_RGT_Hybrids/3_methylation_models/RT_hyb/RT_models/1_RT_model_output_sig_snp_dist.txt", sep = "\t", quote=F, row.names=F)
```

