---
title: "PST-Genetic variation comparison - CPG islands"
author: "Sarah Mueller"
date: "2024-Oct"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
### Libraries
.libPaths(c(.libPaths(),"/dss/dsshome1/lxc0B/ra52qed/R/x86_64-pc-linux-gnu-library/4.2"))

library(vegan)
library(ggplot2)
library(plyr)
library(broom)
library(readxl)
library(cluster)
library(ape)
library(ggrepel)
library(knitr)
library(ggvenn)
library(tidyr)
library(dplyr)
library(ggpmisc)
library(gridExtra)
library(wesanderson)
library(stringr)
```

## PST vs. DXY comparison

DXY calculated on an invariant vcf file (10,000 bp windows) see (/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowWGS_168/1_Population_Structure/4_pixy/HR.168.invariant) then a whole genome average was calculated (see /dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowWGS_168/1_Population_Structure/4_pixy/2_DXY/1_calculate_average_dxy.sh) compared to PST calculated per loci averaged across the whole genome. Here I want to run a correlation between the two. Each point represents one population comparison of PST and genetic variation (FST/DXY).

```{r message=FALSE, warning=FALSE}

#load in just one population comparison first
pst.gw <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/2_Divergence/1_PST/0_cpgi_pst/0_pst_cpgi_c05.txt", header=T)

#sample names
sample_names <- read.table("/dss/dsshome1/lxc0B/ra52qed/gitlab/swallow.projects/scripts/0_sample.lists/sample.names.85.txt")
#add sample pops
sample_names <- sample_names %>%
  mutate(pop = case_when(
    grepl("RU_13_R_", V1) ~ "rustica.r",
    grepl("CH_", V1) ~ "rustica.eu",
    grepl("DE_", V1) ~ "rustica.eu",
    grepl("FI", V1) ~ "rustica.eu",
    grepl("_G_", V1) ~ "gutturalis",
    grepl("_E_", V1) ~ "erythrogaster",
    grepl("_S_", V1) ~ "savignii",
    grepl("_TV_", V1) ~ "transitiva",
    grepl("_T_", V1) ~ "tytleri"))

##Create a list of comparisons
pop.list <- as.data.frame(unique(as.vector(sample_names$pop)))
colnames(pop.list) <- c("pop1")
pop.combo <- tidyr::crossing(var1=pop.list$pop1, var2=pop.list$pop1)
pop.combo <- pop.combo[c(2:7, 10:14, 18:21,26:28, 34:35,42),]

#Get average values across all sites
numeric_columns <- pst.gw[sapply(pst.gw, is.numeric)]

# Calculate the average (mean) of all numeric columns
column_means <- colMeans(numeric_columns, na.rm = TRUE)

# Create a new dataframe with the column names and their corresponding means
mean_df <- data.frame(
  Pops = names(pst.gw[,-1]),
  Mean = column_means
)

mean_df$Pops <- gsub("\\.pst", "", mean_df$Pops)

#load DXY
dxy.gw <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/1_Population_Structure/4_pixy/2_DXY/1_dxy_avg_gw.txt", header=T)
colnames(dxy.gw) <- c("Pop1", "Pop2","avg_dxy")
fst.gw <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/1_Population_Structure/4_pixy/2_FST/1_fst_avg_gw.txt", header=T)
colnames(fst.gw) <- c("Pop1", "Pop2","avg_fst")
fst.islands <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/1_Population_Structure/4_pixy/2_FST/4_fst_cpgi_pop.txt", header=T)
colnames(fst.islands) <- c("Pop1", "Pop2","avg_fst")
fst.opensea <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/1_Population_Structure/4_pixy/2_FST/4_fst_opensea_pop.txt", header=T)
colnames(fst.opensea) <- c("Pop1", "Pop2","avg_fst")
fst.shore <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/1_Population_Structure/4_pixy/2_FST/4_fst_shore_pop.txt", header=T)
colnames(fst.shore) <- c("Pop1", "Pop2","avg_fst")
pi.gw <- read.table("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/5_WGS/1_Population_Structure/4_pixy/2_PI/1_pi_avg_gw.txt", header=T)
colnames(pi.gw) <- c("Pop1", "avg_pi")

#pop key function
generate_pop_key <- function(pop1, pop2) {
  sorted_pops <- sort(c(as.character(pop1), as.character(pop2)))
  return(paste(sorted_pops[1], sorted_pops[2], sep = "_"))
}

#add pop_key to all genetic dataframes
dxy.gw <- dxy.gw %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))
fst.gw <- fst.gw %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))
fst.islands <- fst.islands %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))
fst.opensea <- fst.opensea %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))
fst.shore <- fst.shore %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

```

Next we can load in the homer files, as we may also want to explore the relationships between the different genomic annotations
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Dataset 1: 
#import homer output
homer.cpgi.prop.20miss <- read.delim("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/0_Pre-processing/4_methyl_calling/1_CpG_islands/168.20miss/homer.cpgi.prop.168.merged.20miss.rmlowv.rmout.txt", sep = "\t", header=T)

#join pst and homer
joined.20miss <- right_join(pst.gw, homer.cpgi.prop.20miss, by = c("site.id" = "region.id"))
summary(as.factor(joined.20miss$Annotation))

joined.20miss$Annotation <- trimws(joined.20miss$Annotation)
joined.20miss$Annotation <- trimws(joined.20miss$category)

#remove trailing spaces
joined.20miss <- joined.20miss %>%
  mutate(category = str_trim(category, side = "right"))

# Calculate means for the opensea dataframe
opensea.20miss <- joined.20miss %>%
  filter(grepl("opensea", category))

opensea_column_means <- colMeans(opensea.20miss[, 2:22], na.rm = TRUE)

opensea_mean_df <- data.frame(
  Pops = names(opensea.20miss[, 2:22]),
  Mean = opensea_column_means
)
opensea_mean_df$Pops <- gsub("\\.pst", "", opensea_mean_df$Pops)

# Calculate means for the island dataframe
island.20miss <- joined.20miss %>%
  filter(grepl("island", category))

island_column_means <- colMeans(island.20miss[, 2:22], na.rm = TRUE)

island_mean_df <- data.frame(
  Pops = names(island.20miss[, 2:22]),
  Mean = island_column_means
)
island_mean_df$Pops <- gsub("\\.pst", "", island_mean_df$Pops)

# Calculate means for the shore dataframe
shore.20miss <- joined.20miss %>%
  filter(grepl("shore", category))

shore_column_means <- colMeans(shore.20miss[, 2:22], na.rm = TRUE)

shore_mean_df <- data.frame(
  Pops = names(shore.20miss[, 2:22]),
  Mean = shore_column_means
)
shore_mean_df$Pops <- gsub("\\.pst", "", shore_mean_df$Pops)

```

Save output 
```{r}
#make this table 
#write.table(joined.20miss, file="/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/2_Divergence/1_PST_FST_gw/1_pst_homer/1_pst_homer_cpgi_85_20miss.txt", sep="\t", quote=F, row.names=F)
```
## For only islands:
```{r}
pst_columns <- grep("\\.pst$", colnames(island.20miss), value = TRUE)

# Calculate the average pst values per unique entry in the 'category' column
average_pst_by_category <- island.20miss %>%
  dplyr::select(category, all_of(pst_columns)) %>%  # Select 'category' and pst columns
  group_by(category) %>%                     # Group by 'category'
  summarise(across(everything(), ~mean(.x, na.rm = TRUE)))  # Calculate mean using an anonymous function

# View the result
head(average_pst_by_category)

long_df <- average_pst_by_category %>%
  pivot_longer(
    cols = ends_with(".pst"), # Modify this if needed
    names_to = "population_comparison",
    values_to = "pst_value"
  )

# Create the boxplot
ggplot(long_df, aes(x = category, y = pst_value, fill = category)) +
  geom_boxplot() +
  labs(x = "Category", y = "PST Values", title = "Boxplot of PST Values by Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Define colors for specific comparisons
custom_colors <- c(
  "erythrogaster.vs.gutturalis.pst" = "black",
  "rustica.eu.vs.gutturalis.pst" = "grey66",
  "rustica.a.vs.tytleri.pst" = "grey",
  "gutturalis.vs.rustica.eu.pst" = "brown",
  "gutturalis.vs.transitiva.pst" = "tan"
)

# Get unique population comparisons
unique_comparisons <- unique(long_df$population_comparison)

# Generate rainbow colors for the remaining comparisons
remaining_comparisons <- setdiff(unique_comparisons, names(custom_colors))
rainbow_colors <- rainbow(length(remaining_comparisons))

# Combine custom and rainbow colors
final_colors <- c(custom_colors, setNames(rainbow_colors, remaining_comparisons))

# Create the scatter plot with custom colors
ggplot(long_df, aes(x = category, y = pst_value, color = population_comparison)) +
  geom_point(size = 3) +  # Add jitter for better visibility
  labs(x = "Category", y = "PST Values", title = "PST Values by Category and Population Comparison") +
  theme_minimal() +
  #ylim(0,0.4)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = final_colors) 

```
For DXY calculated from pixy
```{r}
fst.islands <- cbind(fst.islands[,3:4], dxy.gw[,c(1:3)])

#join FST and PST dfs
island_mean_df <- island_mean_df %>%
  separate(Pops, into = c("Pop1", "Pop2"), sep = "\\.vs\\.")

island_mean_df <- island_mean_df %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

merged_df <- merge(fst.islands, island_mean_df, by = "pop_key")

# Create the plot
a <- ggplot(merged_df, aes(x = avg_dxy, y = Mean)) +
  geom_point(size = 3) +
  stat_poly_line() +
  stat_poly_eq() +
  geom_smooth(method = "lm", se = T, color = "grey77", fill = "grey100", fill_alpha = 0.9) +
  labs(title = "Islands",x = "DXY", y = "PST") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.line = element_line(colour = "black"))

cor <- cor(merged_df$Mean, merged_df$avg_dxy, use = "pairwise.complete.obs", method= "spearman")
cor.test(merged_df$Mean, merged_df$avg_dxy, use = "pairwise.complete.obs", method = "spearman")

```

```{r message=FALSE, warning=FALSE}
joined_long <- joined.20miss %>%
  pivot_longer(cols = contains(".vs."), names_to = "Pops", values_to = "Value")

mean_pst <- joined_long %>%
  group_by(Pops, category) %>%
  summarize(mean_PST = mean(Value, na.rm = TRUE)) %>%
  spread(key = category, value = mean_PST)

#remove pst from pops column
mean_pst$Pops <- gsub("\\.pst", "", mean_pst$Pops)

mean_pst <- mean_pst %>%
  separate(Pops, into = c("Pop1", "Pop2"), sep = ".vs.") %>%
  as.data.frame()

mean_pst <- mean_pst %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

# Merge with dxy.gw dataframe
merged_df <- fst.islands %>%
  inner_join(mean_pst, by = c("pop_key"))

#pivot_longer
merged_long <- merged_df %>%
  pivot_longer(cols = -c(Pop1.x, Pop1.y, Pop2.x, Pop2.y, avg_fst, avg_dxy, pop_key), names_to = "Annotation", values_to = "Mean")

#remove a few NAs
merged_islands <- merged_long %>% 
  filter(!Annotation == "<NA>") %>% 
  filter(!grepl("opensea_", Annotation))%>% 
  filter(!grepl("shore_", Annotation))%>% 
  filter(!grepl("shelf_", Annotation))

# Plot
p1 <- ggplot(merged_islands, aes(x = avg_dxy, y = Mean)) +
  geom_point(size = 3) +
  stat_poly_line() +
  stat_poly_eq() +
  #geom_text(data = subset(merged_df, corrected_fst > 0.4), aes(label = site.id), vjust = -0.5, hjust = 1) +
  geom_smooth(method = "lm", se = T, color = "grey77", fill = "grey100", fill_alpha = 0.9) +
  #scale_color_manual(values = comparison_colors) +
  labs(x = "DXY", y = "PST") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  facet_wrap(~ Annotation, drop = TRUE)

p1
```
## For shores:
```{r}
pst_columns <- grep("\\.pst$", colnames(shore.20miss), value = TRUE)

# Calculate the average pst values per unique entry in the 'category' column
average_pst_by_category <- shore.20miss %>%
  dplyr::select(category, all_of(pst_columns)) %>%  # Select 'category' and pst columns
  group_by(category) %>%                     # Group by 'category'
  summarise(across(everything(), ~mean(.x, na.rm = TRUE)))  # Calculate mean using an anonymous function

# View the result
head(average_pst_by_category)

long_df <- average_pst_by_category %>%
  pivot_longer(
    cols = ends_with(".pst"), # Modify this if needed
    names_to = "population_comparison",
    values_to = "pst_value"
  )

# Create the boxplot
ggplot(long_df, aes(x = category, y = pst_value, fill = category)) +
  geom_boxplot() +
  labs(x = "Category", y = "PST Values", title = "Boxplot of PST Values by Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Define colors for specific comparisons
custom_colors <- c(
  "erythrogaster.vs.gutturalis.pst" = "black",
  "rustica.eu.vs.gutturalis.pst" = "grey66",
  "rustica.a.vs.tytleri.pst" = "grey",
  "gutturalis.vs.rustica.eu.pst" = "brown",
  "gutturalis.vs.transitiva.pst" = "tan"
)

# Get unique population comparisons
unique_comparisons <- unique(long_df$population_comparison)

# Generate rainbow colors for the remaining comparisons
remaining_comparisons <- setdiff(unique_comparisons, names(custom_colors))
rainbow_colors <- rainbow(length(remaining_comparisons))

# Combine custom and rainbow colors
final_colors <- c(custom_colors, setNames(rainbow_colors, remaining_comparisons))

# Create the scatter plot with custom colors
ggplot(long_df, aes(x = category, y = pst_value, color = population_comparison)) +
  geom_point(size = 3) +  # Add jitter for better visibility
  labs(x = "Category", y = "PST Values", title = "PST Values by Category and Population Comparison") +
  theme_minimal() +
  #ylim(0,0.4)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = final_colors) 

```
For DXY calculated from pixy
```{r}
fst.shore <- cbind(fst.shore[,3:4], dxy.gw[,c(1:3)])

#join FST and PST dfs
shore_mean_df <- shore_mean_df %>%
  separate(Pops, into = c("Pop1", "Pop2"), sep = "\\.vs\\.")

shore_mean_df <- shore_mean_df %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

merged_df <- merge(fst.shore, shore_mean_df, by = "pop_key")

# Create the plot
b <- ggplot(merged_df, aes(x = avg_dxy, y = Mean)) +
  geom_point(size = 3) +
  stat_poly_line() +
  stat_poly_eq() +
  geom_smooth(method = "lm", se = T, color = "grey77", fill = "grey100", fill_alpha = 0.9) +
  labs(title = "Flanks",x = "DXY", y = "PST") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.line = element_line(colour = "black"))


cor <- cor(merged_df$Mean, merged_df$avg_dxy, use = "pairwise.complete.obs", method= "spearman")
cor.test(merged_df$Mean, merged_df$avg_dxy, use = "pairwise.complete.obs", method = "spearman")
```

```{r message=FALSE, warning=FALSE}
joined_long <- joined.20miss %>%
  pivot_longer(cols = contains(".vs."), names_to = "Pops", values_to = "Value")

mean_pst <- joined_long %>%
  group_by(Pops, category) %>%
  summarize(mean_PST = mean(Value, na.rm = TRUE)) %>%
  spread(key = category, value = mean_PST)

#remove pst from pops column
mean_pst$Pops <- gsub("\\.pst", "", mean_pst$Pops)

mean_pst <- mean_pst %>%
  separate(Pops, into = c("Pop1", "Pop2"), sep = ".vs.") %>%
  as.data.frame()

mean_pst <- mean_pst %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

# Merge with dxy.gw dataframe
merged_df <- fst.shore %>%
  inner_join(mean_pst, by = c("pop_key"))

#pivot_longer
merged_long <- merged_df %>%
  pivot_longer(cols = -c(Pop1.x, Pop1.y, Pop2.x, Pop2.y, avg_fst, avg_dxy, pop_key), names_to = "Annotation", values_to = "Mean")

#remove a few NAs
merged_shore <- merged_long %>% 
  filter(!Annotation == "<NA>") %>% 
  filter(!grepl("island_", Annotation))%>% 
  filter(!grepl("opensea_", Annotation))%>% 
  filter(!grepl("shelf_", Annotation))

# Plot
p1 <- ggplot(merged_shore, aes(x = avg_dxy, y = Mean)) +
  geom_point(size = 3) +
  stat_poly_line() +
  stat_poly_eq() +
  #geom_text(data = subset(merged_df, corrected_fst > 0.4), aes(label = site.id), vjust = -0.5, hjust = 1) +
  geom_smooth(method = "lm", se = T, color = "grey77", fill = "grey100", fill_alpha = 0.9) +
  #scale_color_manual(values = comparison_colors) +
  labs(x = "DXY", y = "PST") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  facet_wrap(~ Annotation, drop = TRUE)

p1
```


## For opensea:

```{r}
pst_columns <- grep("\\.pst$", colnames(opensea.20miss), value = TRUE)

# Calculate the average pst values per unique entry in the 'category' column
average_pst_by_category <- opensea.20miss %>%
  dplyr::select(category, all_of(pst_columns)) %>%  # Select 'category' and pst columns
  group_by(category) %>%                     # Group by 'category'
  summarise(across(everything(), ~mean(.x, na.rm = TRUE)))  # Calculate mean using an anonymous function

# View the result
head(average_pst_by_category)

long_df <- average_pst_by_category %>%
  pivot_longer(
    cols = ends_with(".pst"), # Modify this if needed
    names_to = "population_comparison",
    values_to = "pst_value"
  )

# Create the boxplot
ggplot(long_df, aes(x = category, y = pst_value, fill = category)) +
  geom_boxplot() +
  labs(x = "Category", y = "PST Values", title = "Boxplot of PST Values by Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Define colors for specific comparisons
custom_colors <- c(
  "erythrogaster.vs.gutturalis.pst" = "black",
  "rustica.eu.vs.gutturalis.pst" = "grey66",
  "rustica.a.vs.tytleri.pst" = "grey",
  "gutturalis.vs.rustica.eu.pst" = "brown",
  "gutturalis.vs.transitiva.pst" = "tan"
)

# Get unique population comparisons
unique_comparisons <- unique(long_df$population_comparison)

# Generate rainbow colors for the remaining comparisons
remaining_comparisons <- setdiff(unique_comparisons, names(custom_colors))
rainbow_colors <- rainbow(length(remaining_comparisons))

# Combine custom and rainbow colors
final_colors <- c(custom_colors, setNames(rainbow_colors, remaining_comparisons))

# Create the scatter plot with custom colors
ggplot(long_df, aes(x = category, y = pst_value, color = population_comparison)) +
  geom_point(size = 3) +  # Add jitter for better visibility
  labs(x = "Category", y = "PST Values", title = "PST Values by Category and Population Comparison") +
  theme_minimal() +
  #ylim(0,0.4)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_color_manual(values = final_colors) 

```

For DXY calculated from pixy
```{r}
fst.opensea <- cbind(fst.opensea[,3:4], dxy.gw[,c(1:3)])

#join FST and PST dfs
opensea_mean_df <- opensea_mean_df %>%
  separate(Pops, into = c("Pop1", "Pop2"), sep = "\\.vs\\.")

opensea_mean_df <- opensea_mean_df %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

opensea_df <- merge(fst.opensea, opensea_mean_df, by = "pop_key")

# Create the plot
c <- ggplot(opensea_df, aes(x = avg_dxy, y = Mean)) +
  geom_point(size = 3) +
  stat_poly_line() +
  stat_poly_eq() +
  geom_smooth(method = "lm", se = T, color = "grey77", fill = "grey100", fill_alpha = 0.9) +
  labs(title = "Sites",x = "DXY", y = "PST") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        axis.line = element_line(colour = "black"))

cor <- cor(opensea_df$Mean, opensea_df$avg_dxy, use = "pairwise.complete.obs", method= "spearman")
cor.test(opensea_df$Mean, opensea_df$avg_dxy, use = "pairwise.complete.obs", method = "spearman")

```

```{r message=FALSE, warning=FALSE}
joined_long <- joined.20miss %>%
  pivot_longer(cols = contains(".vs."), names_to = "Pops", values_to = "Value")

mean_pst <- joined_long %>%
  group_by(Pops, category) %>%
  summarize(mean_PST = mean(Value, na.rm = TRUE)) %>%
  spread(key = category, value = mean_PST)

#remove pst from pops column
mean_pst$Pops <- gsub("\\.pst", "", mean_pst$Pops)

mean_pst <- mean_pst %>%
  separate(Pops, into = c("Pop1", "Pop2"), sep = ".vs.") %>%
  as.data.frame()

mean_pst <- mean_pst %>%
  rowwise() %>%
  mutate(pop_key = generate_pop_key(Pop1, Pop2))

# Merge with dxy.gw dataframe
merged_df <- fst.opensea %>%
  inner_join(mean_pst, by = c("pop_key"))

#pivot_longer
merged_long <- merged_df %>%
  pivot_longer(cols = -c(Pop1.x, Pop1.y, Pop2.x, Pop2.y, avg_fst, avg_dxy, pop_key), names_to = "Annotation", values_to = "Mean")

#remove a few NAs
merged_opensea <- merged_long %>% 
  filter(!Annotation == "<NA>") %>% 
  filter(!grepl("island_", Annotation))%>% 
  filter(!grepl("shore_", Annotation))%>% 
  filter(!grepl("shelf_", Annotation))

# Plot
p3 <- ggplot(merged_opensea, aes(x = avg_dxy, y = Mean)) +
  geom_point(size = 3) +
  stat_poly_line() +
  stat_poly_eq() +
  #geom_text(data = subset(merged_df, corrected_fst > 0.4), aes(label = site.id), vjust = -0.5, hjust = 1) +
  geom_smooth(method = "lm", se = T, color = "grey77", fill = "grey100", fill_alpha = 0.9) +
  #scale_color_manual(values = comparison_colors) +
  labs(x = "DXY", y = "PST") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  facet_wrap(~ Annotation, drop = TRUE)

p3
```

```{r}
# Arrange the three plots in 1 row and 3 columns
grid.arrange(a, b, c, ncol = 3)

```
save results
```{r}
ggsave("/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/2_Divergence/1_PST_FST_gw/SuppFig_PST_DXY_regional.png", arrangeGrob(a, b, c, ncol = 3), width = 12, height = 4, units = "in", dpi = 500)


all_regions_pst <- rbind(merged_islands, merged_shore, merged_opensea)

write.table(all_regions_pst, "/dss/dsslegfs01/pr53da/pr53da-dss-0034/projects/2021_SwallowRRBS/2_Divergence/1_PST_FST_gw/1_rho_dxy_per_category_results.txt", quote=F, row.names=F)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
